Progress Report

### 2023-09-20 Docs/ProgressReport.Rmd
investigate effect of linear interpolation

- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
- 3.3 minutes spent on model fitting, 20 minutes spent on prediction (about 0.01 per subject)
- Three subjects had failed approximation, all happened with the shortest observed track

@@

- Example of out-of-sample prediction

```{r message=FALSE, warning=FALSE}
rand_id2 <- sample(unique(df_test$id), 4)
df %>% 
  filter(id %in% rand_id2) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020)) %>%
  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>% 
  ggplot()+
    geom_line(aes(x=bin, y = pred_t540, col = "9am"), linetype = "dashed")+
    geom_line(aes(x=bin, y = pred_t780, col = "1pm"),linetype = "dashed")+
    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"),linetype = "dashed")+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

- Subjects with numeric issue

```{r}
df %>% filter(id %in% skip_id) %>% 
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.2)+
  facet_wrap(~id)+
  geom_vline(xintercept = c(540))
```

@@

Since we are assuming value of latent function is constant within a bin, why can't we just do the same when calculating AUC without interpolation?

Remove the three subjects with failed approximation

```{r}
lst_auc <- df %>% 
  filter(id %in% unique(df_test$id) & !id %in% skip_id ) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin")) %>% 
  mutate(window = cut(sind, breaks = c(0, 540, 780, 1020, 1260, 1440), 
                      labels = c("0-9am", "9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
lst_auc <- split(lst_auc, f=lst_auc$window)

# up to 9am
auc_t540 <- lst_auc[2:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t540, x$Y), measure = "auc"))
auc_t540 <- lapply(auc_t540, function(x){x@y.values[[1]]}) %>% unlist()

@@

@@

```{r}
options(knitr.kable.NA = '')
tb_auc <- data.frame(auc_t540, 
           c(NA, auc_t780), 
           c(NA, NA, auc_t1020), 
           auc_t540_adglmm,
           c(NA, auc_t780_adglmm), 
           c(NA, NA, auc_t1020_adglmm))

colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 2)

tb_auc %>%
  kable(digit = 2) %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c("Time spent on model fitting" = 1, "3.3" = 3, "20.72" = 3), bold = F) %>%
  add_header_above(c("Time spent on prediction" = 1, "20.0" = 3, "12.2" = 3), bold = F) %>% 
  add_header_above(c(" "=1, "fGPFCA"=3, "GLMMadaptive"=3))
```

@@

- What would be the reference method to compare our performance to? Original GLMMadaptvie perhaps is feasible on teh entire dataset. But what should we use if we wanna do sub-sample comparison? Should we even do that for this data application (or simulation alone)? 
- There are still three folks with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects.

>>>>

- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
- 3.3 minutes spent on model fitting, 20 minutes spent on prediction (about 0.01 per subject)
- Three subjects had failed approximation, all happened with the shortest observed track
- We can either assume that the latent function value is constant within the same bin, or use linear interpolation to fill in the gaps between bin midpoints. Here I wanna compare if this choice would make any different. Also, there are a log of other methods to fill in the gaps, such as projection. 
- For linear interpolation, I set the edges to be the closest data extreme.

@@

```{r}
# linear interpolation
# remove subjects with numeric problem
df_test_full <- df %>% 
  filter(id %in% unique(df_test$id)) %>%
  filter(!id %in% skip_id) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))
df_test_full$pred_t540_interp <- df_test_full$pred_t780_interp <- df_test_full$pred_t1020_interp <- NA

# interpolation for each subject
for(i in unique(df_test_full$id)){
  # interpolation t540
  interp_i_t540 <- approx(x = df_test[df_test$id==i, "bin"], 
                     y = df_test[df_test$id==i, "pred_t540"],
                     xout = 541:J, rule = 2)
  df_test_full$pred_t540_interp[df_test_full$id==i & df_test_full$sind>540] <- interp_i_t540$y

  # interpolation t780
  interp_i_t780 <- approx(x = df_test[df_test$id==i, "bin"], 
                     y = df_test[df_test$id==i, "pred_t780"],
                     xout = 781:J, rule = 2)
  df_test_full$pred_t780_interp[df_test_full$id==i & df_test_full$sind>780] <- interp_i_t780$y

  # interpolation t1020
  interp_i_t1020 <- approx(x = df_test[df_test$id==i, "bin"], 
                     y = df_test[df_test$id==i, "pred_t1020"],
                     xout = 1021:J, rule = 2)
  df_test_full$pred_t1020_interp[df_test_full$id==i & df_test_full$sind>1020] <- interp_i_t1020$y
}

Let's look at four subjects as an example:

```{r message=FALSE, warning=FALSE}
rand_id2 <- sample(unique(df_test$id), 4)
# without interpolation, assume constant latent function value in each bin
df_test_full%>% 
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>% 
  ggplot()+
    geom_line(aes(x=bin, y = pred_t540, col = "9am"))+
    geom_line(aes(x=bin, y = pred_t780, col = "1pm"))+
    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"))+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)", 
         title = "Constant latent function within bin")
```

```{r, warning=FALSE}
# with interpolation
df_test_full%>% 
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t540_interp, pred_t780_interp, pred_t1020_interp), function(x)exp(x)/(1+exp(x))) %>% 
  ggplot()+
    geom_line(aes(x=bin, y = pred_t540_interp, col = "9am"))+
    geom_line(aes(x=bin, y = pred_t780_interp, col = "1pm"))+
    geom_line(aes(x=bin, y = pred_t1020_interp, col = "5pm"))+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)", 
         title = "Interpolated latent function")

```
- Subjects with numeric issue

```{r, fig.height=3, fig.width=9}
df %>% filter(id %in% skip_id) %>% 
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.2)+
  facet_wrap(~id)+
  geom_vline(xintercept = c(540))
```

@@

```{r}
# break by prediction window
lst_auc <- df_test_full %>% 
  mutate(window = cut(sind, breaks = c(0, 540, 780, 1020, 1260, 1440), 
                      labels = c("0-9am", "9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
lst_auc <- split(lst_auc, f=lst_auc$window)
```

```{r}
# no interpolation
# up to 9am
auc_t540 <- lst_auc[2:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t540, x$Y), measure = "auc"))
auc_t540 <- lapply(auc_t540, function(x){x@y.values[[1]]}) %>% unlist()

@@

```{r}
# with interpolation
# up to 9am
auc_t540_interp <- lst_auc[2:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t540_interp, x$Y), measure = "auc"))
auc_t540_interp <- lapply(auc_t540_interp, function(x){x@y.values[[1]]}) %>% unlist()

# up to 1pm
auc_t780_interp <- lst_auc[3:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t780_interp, x$Y), measure = "auc"))
auc_t780_interp <- lapply(auc_t780_interp, function(x){x@y.values[[1]]}) %>% unlist()

# up to 5pm
auc_t1020_interp <- lst_auc[4] %>% 
  lapply(function(x)performance(prediction(x$pred_t1020_interp, x$Y), measure = "auc"))
auc_t1020_interp <- lapply(auc_t1020_interp, function(x){x@y.values[[1]]}) %>% unlist()
```

@@

```{r}
options(knitr.kable.NA = '')
tb_auc <- data.frame(auc_t540, 
           c(NA, auc_t780), 
           c(NA, NA, auc_t1020), 
           auc_t540_interp, 
           c(NA, auc_t780_interp), 
           c(NA, NA, auc_t1020_interp), 
           auc_t540_adglmm,
           c(NA, auc_t780_adglmm), 
           c(NA, NA, auc_t1020_adglmm))

colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 3)

tb_auc %>%
  kable(digit = 4) %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c("Time spent on model fitting" = 1, "3.3" = 6, "20.72" = 3), bold = F) %>%
  add_header_above(c("Time spent on prediction" = 1, "20.0" = 6, "12.2" = 3), bold = F) %>% 
  add_header_above(c(" "=1, "fGFPCA"=3, "fGFPCA+interpolation" = 3, "GLMMadaptive"=3))
```

@@

- What would be the reference method to compare our performance to? Original GLMMadaptvie perhaps is feasible on teh entire dataset. But what should we use if we wanna do sub-sample comparison? Should we even do that for this data application (or simulation alone)? 
- There are still three folks with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects. 
- Is interplation really necessary? It does not improve performance very much but consumes time. Also, should I try more sophisticated methods for grid extension?
